---
title: "Clovis Allometry"
author: "David K. Thulman"
date: "5/15/2020"
output: github_document
---
This R code creates most of the figures and tables in the Clovis Point Allometry: Exploring Shape Variation with Landmark-based Geometric Morphometrics. 

It uses two packages beyond the basic R packages: (1) geomorph and (2) Hmisc. To start, load those packages.

```{r}
library(geomorph)
library(Hmisc)
```
Now, set the working directory to the GitHub repository: https://github.com/dktArcoop/Clovis-Allometry-Files

```{r}
setwd(getwd())
```

The first step is to set up the background files in geomorph. These first files will be the tps file for the 23 Landmark (LM) configuration, which is aligned with a GPA and converted into a geomorph dataframe. Second, a csv file will be uploaded and appended to the dataframe. The natural logs for several measurements will created in the dataframe. 

```{r}
CAP23<-readland.tps("CAP23.tps", specID = "ID") #creates new file with 23LM tps data

Z.gpa<-gpagen(CAP23) #Procrustes alignment renamed as Z.gpa

groups<-read.csv("APF Cache.csv",header=TRUE,stringsAsFactors = TRUE) #reads in the classifier data

CAP23.gdf <- geomorph.data.frame(Z.gpa) #creates a dataframe from the 23LM gpagen output

CAP23.gdf<-append(CAP23.gdf, groups) #appends groups factors to the dataframe created from Z.gpa, keeps same name

CAP23.gdf$lnCZE <- log(CAP23.gdf$Csize) #Creates new column with log of Centroid Size
CAP23.gdf$lnCZS <- log(CAP23.gdf$StemCZ) #Creates new column with log of Blade centroid Size
CAP23.gdf$lnCZB <- log(CAP23.gdf$BladeCZ) #Creates new column with log of Stem Centroid Size
```
Now, the Mean Shape of the data will be created, along with a links file that connects the LMs. The links file is created with an interactive function and has been disabled. A previously created links file is included.
```{r}
MS <- mshape(CAP23.gdf$coords) #creates mean shape and labels LMs 
par(mfcol = c(1,1))#sets single-plot option
#EMSL<- define.links(MS) #create a links file named EMSL. Interactive function disabled
EMSL <- read.csv("EMSL.csv")
CAP23.gdf<-append(CAP23.gdf, EMSL) #append the links file EMSL to the dataframe
plot(MS, links=EMSL) #Plots mean shape [Figure 2A]
```

A plot of the distribution of aligned LMs from the 100 Clovis points is created.

```{r}
plotAllSpecimens(CAP23.gdf$coords, mean=TRUE, plot.param = list(pt.cex=.6), links=EMSL) #Distibution of LMs from all specimens with the links file [Figure 3A]
```

This plots the reference shape (the mean shape) to one of the larger Clovis points to illustrate the vector displacement of LMs.

```{r}
plotRefToTarget(MS, CAP23.gdf$coords[,,11], links=EMSL, method="vector") #vector displacement from mean to specimen 11 [Figure 6A]
#mtext("Vector Displacement") #title disabled
```

This creates a PCA with wiregrid illustrations of points on the max and min points of PCA1. This was not a figure in the paper.

```{r}
plotTangentSpace(CAP23.gdf$coords, groups=CAP23.gdf$Cache)#creates PCA plot with max and min images on PC1, colors for non-cacke (red) and cache (black)
```

The Allometry Analysis. This creates Figure 5.

```{r}
attach(CAP23.gdf)
color = rep(NA, length=length(CAP23.gdf$Cache)) #colors cache and non-cache dots in plot
color[which(CAP23.gdf$Cache=="Cache")] = "black"
color[which(CAP23.gdf$Cache=="Other")] = "red"
color #check to determine color designations are correct

CAP23CS <- procD.lm(coords~ log(Csize), data=CAP23.gdf, iter=10000, RRPP=TRUE, print.progress = FALSE) #names regression of coords on log Centroid size
plotAllometry(CAP23CS, size=CAP23.gdf$lnCZE,  method = "PredLine", col=color, pch=19, bg=CAP23.gdf$Cache, xlab="Ln Centroid Size") #plots the Prediction Line
CAPlnCS <- plotAllometry(CAP23CS, size=CAP23.gdf$lnCZE,  method = "PredLine", col=color, pch=19, bg=CAP23.gdf$Cache, xlab="Ln Centroid Size") #plots the Prediction Line
#picknplot.shape(CAPlnCS) #creates the small inserts for Figure 4, interaction function disabled
```

Creates the ANOVA results in Table 2 illustrating the presence of allometry and amount of change (r2) for log normal of the entire, stem, and base centroid sizes. 

```{r}

CAP23CS.anova <- anova(CAP23CS) #calculates the anova of the Clovis Allometry on log Centroid Size. CAP23CS is linear regression object created above
summary(CAP23CS.anova) #summary of the ANOVA results against the null of isometric change

CAP23SCS <- procD.lm(coords~ lnCZS, data=CAP23.gdf, iter=10000, RRPP=TRUE, print.progress = FALSE) #regression on of entire shape on log stem centroid size. Stem CS calculated below and then included in the "groups.csv" 
CAP23SCS.anova <- anova(CAP23SCS) #calculates the anova 
summary(CAP23SCS.anova) #summary of the ANOVA results against the null of isometric change

CAP23BCS <- procD.lm(coords~ lnCZB, data=CAP23.gdf, iter=10000, RRPP=TRUE, print.progress = FALSE) ##regression on of entire shape on log stem centroid size. Blade CS calculated below and then included in the "groups.csv"
CAP23BCS.anova <- anova(CAP23BCS) #calculates the anova
summary(CAP23BCS.anova) #summary of the ANOVA results against the null of isometric change
```

The Integration Test

```{r}
CAP23.inter <- integration.test(Z.gpa$coords[c(1:4,6:13),,], Z.gpa$coords[c(5,14:23),,], iter=999, print.progress = FALSE) # divide entire shape into stem and blade modules
summary(CAP23.inter) #provides r-PLS score
plot(CAP23.inter) #plots PLS Blocks 1 and 2, not included as a figure
```

The Modularity Test and Figure 7

```{r}
CAP.gps <- c("A","A","A","A","B","A","A","A","A","A","A","A","A","B","B","B","B","B","B","B","B","B","B") # makes stem(A) and Blade(B) modules
modularity.test(Z.gpa$coords, CAP.gps, iter=999, print.progress = FALSE) #calculates CR statistic
CAP23.mod <- modularity.test(Z.gpa$coords, CAP.gps, iter=999, print.progress = FALSE) #creates object for histogram

hist(CAP23.mod$random.CR,xlim = c(0.96, 1.09), xlab="CR Coefficient", main="CR Coefficient Distribution", col="gray") #histogram showing CR results. The "hist" function allows scale changes on X axis
abline(v = 1.0064,col="red") #abcissa line [Figure 7]
```

Analyses on just the Clovis Point Stems using the non-cache points

Preliminaries

```{r}
#The following analyses concern the Clovis stems. The stem shape is excerpted from the overall shape after GPA by deleting the LMs making up the blade shape.

CAPS12Other<-readland.tps("CAP23 NonCache.tps", specID = "ID") #creates new file with 23LM tps Non-Cache data

CAPSt12O.gpa<-gpagen(CAPS12Other) #Procrustes alignment renamed as CAP23O.gpa

groupsNC<-read.csv("APF Cache Other.csv",header=TRUE,stringsAsFactors = TRUE) #reads in the same classifier data with 75 non-cache points

CAPST12O.gdf<- geomorph.data.frame(CAPSt12O.gpa) #creates a dataframe from the 23LM gpagen output

CAPST12O.gdf<-append(CAPST12O.gdf, groupsNC) #appends groups factors to the dataframe created from CAP23.gfa, keeps same name

CAPST12O.gdf$lnSCS <- log(CAPST12O.gdf$Csize) #create new column with natural log of entire point centroid size

omit <- c(5,14:23)# creates file of LMs (named "omit") to delete the blade LMs from the aligned entire shape
SS<-CAPST12O.gdf$coords[-omit,,]# omits blade LMs coords list and creates stem file SS
CAPST12O.gdf$coords<-SS #replaces original coords file with stem file SS

CAPST12O.gdf$lnCZS <- log(CAPST12O.gdf$StemCZ)#create new column with natural log of entire point centroid size
```

Creating the Mean Shape of the Stem (not used as a figure)

```{r}
MeanStem <- mshape(CAPST12O.gdf$coords) #creates object from meanshape of stem
#MSL<-define.links(MeanStem) #interactive funtion (disabled) creates links file 
MSL <- read.csv("MSL.csv")
CAPST12O.gdf<-append(CAPST12O.gdf, MSL) #appends links file MSL
plot(MeanStem, links=MSL) #plot LM numbers on mean shape with links
```

Plot of the distribution of 75 non-cache Clovis points around the stem LMs (not used as a figure)

```{r}
plotAllSpecimens(CAPST12O.gdf$coords, mean=TRUE, links=MSL, plot.param = list(pt.cex=.6))#plots distribution of 75 points on the mean stem shape   
```

Vector displacement on just the stem to large Clovis point (Figure 6B)

```{r}
plotRefToTarget(MeanStem, CAPST12O.gdf$coords[,,11], links=MSL, method="vector") #vector displacement from mean to specimen 11, a large Clovis point [Figure 7B]
#mtext("Vector Displacement") #plot title disabled
```

The ANOVA results for stem shape by size-class (table 4)

```{r}
fitSO <- procD.lm(coords ~ ClassO, data = CAPST12O.gdf, print.progress = FALSE) #Comparison of Stem shape by size-class
anova(fitSO) #Results show no difference among size classes
```

The ANOVA results for stem shape on log normal for stem centroid size, width, and length (table 4)

```{r}
fitSOCZ <- procD.lm(coords ~ lnCZS, data = CAPST12O.gdf, print.progress = FALSE) #regression of stem shape on ln stem centroid size
anova(fitSOCZ) #Results show significant differences 

fitSASW <-  procD.lm(coords~ log(StemW), data=CAPST12O.gdf, iter=999, RRPP=TRUE, print.progress = FALSE) #regression on ln stem width
anova(fitSASW)

fitSASL <-  procD.lm(coords~ log(StemL), data=CAPST12O.gdf, iter=999, RRPP=TRUE, print.progress = FALSE) #regression on log stem length
anova(fitSASL)
```

Bivariate plot of Stem Width and Stem Length (figure 8)

```{r}
attach(CAPST12O.gdf)

par(mfcol = c(1,1))
#Figure 7
plot(groups[,3:4],xaxt="none",yaxt="none", xlab="", ylab="")#sets plot box and 75 points
axis(1,seq(0,50,5)) #defines x-axis
axis(2,seq(0,40,5),las=2) #defines y-axis
#mtext(side=3, line=0.2,"Scatterplot Stem Width and Length", font=2, cex=1.5) #plot title, disabled
minor.tick(nx = 10, ny = 5, tick.ratio=0.5) #minor ticks on axis, size and number
title(ylab="Stem Width", line=2.25) #y-axis label, size and position
title(xlab="Stem Length",line=2.25) #x-axis label, size and position
rect(10.5,12.5,26.5,18.5) # lower rectangle
rect(12.5,19.5,37.5,26.5) # middle rectangle
rect(19.5,26.5,46.5,34) # upper rectangle
points(groups[which(ClassO2=="S"),3:4],pch=21,cex=1,col="blue",bg="blue") #color and shape of small size class
points(groups[which(ClassO2=="M"),3:4],pch=22,cex=1,col="green",bg="green") #color and shape of medium size class
points(groups[which(ClassO2=="L"),3:4],pch=24,cex=1,col="red",bg="red") #color and shape of large size class
points(groups[which(ClassO2=="BWM"),3:4],pch=5,cex=1,col="black") #shapes for small BWD points
points(groups[which(ClassO2=="CBY"),3:4],pch=3,cex=1,col="black") #shape for large Colby point
```

Bivariate plot of Stem Width and Blade Length (figure 9)

```{r}
plot(groups[,5:4],xaxt="none",yaxt="none", xlab="", ylab="")
axis(1,seq(15,90,5))
axis(2,seq(0,40,5),las=2)
#mtext(side=3, line=0.2,"Scatterplot Stem Width and Blade Length", font=2, cex=1.5)
minor.tick(nx = 10, ny = 5, tick.ratio=0.5)
title(ylab="Stem Width", line=2.25)
title(xlab="Blade Length",line=2.25)
rect(18.5,12.5,49,18.5)
rect(21.5,19.5,89,26.5)
rect(43.5,26.5,89,34)
points(groups[which(ClassO2=="S"),5:4],pch=21,cex=1,col="blue",bg="blue")
points(groups[which(ClassO2=="M"),5:4],pch=22,cex=1,col="green",bg="green")
points(groups[which(ClassO2=="L"),5:4],pch=24,cex=1,col="red",bg="red")
points(groups[which(ClassO2=="BWM"),5:4],pch=5,cex=1,col="black")
points(groups[which(ClassO2=="CBY"),5:4],pch=3,cex=1,col="black")
```

Bivariate plot of Stem Length and Blade Length (figure 10)

```{r}
plot(groups[,c(5,3)],xaxt="none",yaxt="none", xlab="", ylab="")
axis(1,seq(15,90,5))
axis(2,seq(0,50,5),las=2)
#mtext(side=3, line=0.2,"Scatterplot Stem Length and Blade Length", font=2, cex=1.5)
minor.tick(nx = 10, ny = 10, tick.ratio=0.5)
title(ylab="Stem Length", line=2.25)
title(xlab="Blade Length",line=2.25)
rect(18.5,10.25,44.5,19.5)
rect(28.25,19.5,89,46.75)
points(groups[which(ClassO2=="S"),c(5,3)],pch=21,cex=1,col="blue",bg="blue")
points(groups[which(ClassO2=="M"),c(5,3)],pch=22,cex=1,col="green",bg="green")
points(groups[which(ClassO2=="L"),c(5,3)],pch=24,cex=1,col="red",bg="red")
points(groups[which(ClassO2=="BWM"),c(5,3)],pch=5,cex=1,col="black")
points(groups[which(ClassO2=="CBY"),c(5,3)],pch=3,cex=1,col="black")
```

Generating Stem and Blade Centroid Sizes

```{r}
CAPSCZ<-readland.tps("CAPStem12.tps", specID = "ID") #creates new file with 12LM tps for stem. tps file created un tpsUtil.
CAP12SCZ<-gpagen(CAPSCZ) #Procrustes alignment renamed as CAP12SCZ.gpa
CAP12SCZ$Csize #list of stem centroid sizes. List included in groups csv file


CAPBCZ<-readland.tps("CAPBlade11.tps", specID = "ID") #creates new file with 12LM tps for stem. tps file created un tpsUtil.
CAP11BCZ<-gpagen(CAPBCZ) #Procrustes alignment renamed as CAP12SCZ.gpa
CAP11BCZ$Csize #list of blade centroid sizes. List included in groups csv file
```

END
